ASTR480 Final Report:\\ \textit{Asteroids in TESS}

### 2024-09-11 FinalReport/FinalReportDraft.tex
more writting at ISO chat

\DeclareSIUnit{\arcsec}{''}
\DeclareSIUnit{\au}{\rm{AU}}
\DeclareSIUnit{\mag}{\rm{mag}}

@@

This work aims to extend these studies to more sectors, hence characterising more asteroids.
The aim is to get a volume complete set of asteroid lightcurves and periods, by analysis of the whole sky as seen by TESS using data from more than just the first year of TESS's operation.
I also employ a different data reduction method, reducing the data with the \texttt{TESSreduce} package \citep{Ridden-Harper2021} should increase the accuracy of the photometry due to the improved background subtraction process.
Because of the survey properties, TESS provides a self-consistent way to measure the properties of asteroids over the full sky.
By using the higher cadence FFIs of later TESS sectors, the $P\leq\qty{3}{\hour}$ limit on the accurate periods \citep[as found by ][]{McNeill2023} should be lowered.  
Another beneficial part of this work is that as part of a full sky transient survey using TESS, \texttt{TESSELLATE} \citep{TESSELATE}.
Asteroids can spike the brightness of a pixel for only a few frames while it is passing by.
This can confuse pipelines looking for  transients such as flare stars and supernova, as they can look very similar, having the brightness change in one pixel for a short time.
The goal of finding all the asteroids will allow for the removal of these spikes from the transient pipeline, as well as to understand the asteroid population better.
By filtering the asteroid detections out, while also self consistently determining rotation properties of a large sample of the population, more science can be done with the one action.

@@

With TESS data coming in $\frac12\unit{\hour}$ chunks, 24 interpolated points are needed between each API call.
This interpolation should be accurate, as asteroids move at close to a TESS pixel per FFI on average \citep{Pal2018}, and checking against a higher frequency query to JPL Horizons confirmed this accuracy.
For the shorter FFIs, the asteroids will be move fewer pixels per frame, which could cause slight mismatches between the interpolation and the detections.
Horizons was also used get properties of the individual asteroids, such as the absolute magnitude $H$.
For the faster TESS data, more interpolated points are needed, but a smaller the change in position between each point.
These interpolated positions can be seen in \autoref{Fig:interpPos} for a cut from \texttt{TESSELLATE}.
There are a few interesting features, such as the asteroids are moving in the same direction, indicated by the colouring, they come in from low RA and Dec and tend to increase both coordinates as the month of the sector progresses.
There is also a large size range in this slice of sky, ranging over \qty{5}{\mag} in absolute magnitude, which can be seen in the alpha (or transparency) of the tracks in \autoref{Fig:interpPos}.

@@

%Match to detections
Matching these interpolated positions to \texttt{TESSELLATE} detections is important to lower the unknown transient outputs of this pipeline.
Having interpolated their positions, the asteroids have a well sampled set of RA, Dec and time values of where they should be in the TESS data.
They should show up in a pixel for a small number of frames, of order $\sim1$.
This is the same as a lot of other transient events, a sharp rise in brightness and then disapering quickly again.
The number of frames do change, type Ia supernova will brighten is a matter of a few hours and then dim for days, while stellar flares are of similar profile by a smaller max brightness and a corrsepondingly shorter decay time.
Asteroids are very short, however detection pipelines are robust.
These pipelines have already found the aforementioned supernova and stellar flares, the job of this work is to catch all the asteroids in the set of all the transients.
To do this, catelogue matching is in order.

@@

%light curves; detected VS forced interpolated 
There are two sets of points to take light curves from.
The matches from the detections, which already have a flux calculated, and the interpolated points themselves, which are more numerous but require forcing the photometry.
There were some challenges getting the flux of the interpolated points, even when \texttt{TESSELLATE} has already reduced all the FFIs of interest, due to the timing of the TESS frames, but these were identified and corrected for.
Not every interpolated point gets a match, due to a variety of reasons, %pass in front of stars, dip below limiting mag, bad pipeline, not extreme enough of a difference between frames 
so a comparison between the two light curves is interesting.
\autoref{Fig:DifFlux} shows these two light curves for a chosen asteroid, ``Ruff''.
This was picked as it has a high number of points matched to detections.
There are more interpolated points than matched points, as not ever interpolated point gets matched to a detection.
There seems to be a systematic offset between the fluxes, with the interpolated points having consistently lower flux, (here the means differ by 58 counts)
Adjusting aperture to sum the fluxes over to the ``centre of mass'' of the asteroid in each frame does not alleviate this problem.

@@

>>>>

\DeclareSIUnit{\arcsec}{''}
\DeclareSIUnit{\au}{\rm{AU}}
\DeclareSIUnit{\mag}{\rm{mag}}
\DeclareSIUnit{\px}{\rm{px}}

@@

This work aims to extend these studies to more sectors, hence characterising more asteroids.
The aim is to get a volume complete set of asteroid lightcurves and periods, by analysis of the whole sky as seen by TESS using data from more than just the first year of TESS's operation.
I also employ a different data reduction method, reducing the data with the \texttt{TESSreduce} package \citep{Ridden-Harper2021} should increase the accuracy of the photometry due to the improved background subtraction process.
Because of the survey properties, TESS provides a self-consistent way to measure the properties of asteroids over the full sky.
By using the higher cadence FFIs of later TESS sectors, the $P\leq\qty{3}{\hour}$ limit on the accurate periods \citep[as found by ][]{McNeill2023} should be lowered.
Another beneficial part of this work is that as part of a full sky transient survey using TESS, \texttt{TESSELLATE} \citep{TESSELLATE}.
Asteroids can spike the brightness of a pixel for only a few frames while it is passing by.
This can confuse pipelines looking for  transients such as flare stars and supernova, as they can look very similar, having the brightness change in one pixel for a short time.
The goal of finding all the asteroids will allow for the removal of these spikes from the transient pipeline, as well as to understand the asteroid population better.
By filtering the asteroid detections out, while also self consistently determining rotation properties of a large sample of the population, more science can be done with the one action.

@@

With TESS data coming in $\frac12\unit{\hour}$ chunks, 24 interpolated points are needed between each API call.
Because of a small change in position between each frame, \citep[$\sim \qty{1}{\px}$ per frame][]{Pal2018}, this interpolation should be accurate.
Checking against a higher frequency query to the JPL Horizons ephemeris  confirmed this accuracy on a few asteroids. %TODO fig
This is not a feasable way of getting all of the positions, as the queries are per name, not per area like the conesearch in Skybot.
They are also rate limited, so while a few asteroids can be checked, filling the gaps bewteen the Skybot half day queries for all the asteroids would take far too long.
For the faster TESS FFIs, more interpolated points are needed, but a smaller the change in position between each point will be.
Forcing the interpolations to be at the same time as the TESS frames, which are known thanks to \texttt{TESSELLATE} saving them out, simplifies the matching to detections and forced photometry of these points.
These interpolated positions can be seen in \autoref{Fig:interpPos} for a cut from \texttt{TESSELLATE}.
There are a few interesting features, such as the asteroids are moving in the same direction, indicated by the colouring, they come in from low RA and Dec and tend to increase both coordinates as the month of the sector progresses.
There is also a large size range in this slice of sky, ranging over \qty{5}{\mag} in absolute magnitude, which can be seen in the alpha (or transparency) of the tracks in \autoref{Fig:interpPos}.

@@

%Match to detections
Matching these interpolated positions to \texttt{TESSELLATE} detections is important to lower the unknown transient outputs of this pipeline.
Having interpolated their positions, the asteroids have a well sampled set of RA, Dec and time values of where they should be in the TESS data.
They should show up in a pixel for a small number of frames, of order $\sim1$.
This is the same as a lot of other transient events, a sharp rise in brightness and then disapering quickly again.
The number of frames do change, type Ia supernova will brighten is a matter of a few hours and then dim for days, while stellar flares are of similar profile by a smaller max brightness and a corrsepondingly shorter decay time.
Asteroids are very short spikes, however these detection pipelines are robust. %? Clarinda begs to differ
These pipelines have already found the aforementioned supernova, stellar flares and variable stars, by matching to catalogues of those types of objects (Gaia, ZTF etc.).
The job of this work is to catch all the asteroids in the set of all the transients that are returned from the pipeline.
To do this, catelogue matching is in order.

@@

%light curves; detected VS forced interpolated 
There are two sets of points to take light curves from.
The matches from the detections, which already have a flux calculated, and the interpolated points themselves, which are more numerous but require forcing the photometry.

There were some challenges getting the flux of the interpolated points, even when \texttt{TESSELLATE} has already reduced all the FFIs of interest.
The flux is summed in  $3\times3$ box centred on the integer $xy$ position returned from changing the world coordinates in RA and Dec to pixel coordinates using \texttt{Astropy}'s \texttt{SkyCoord} subpackage. %? cite

The integer coordinates of the asteroids seemed to be leading the object in any given frame of TESS. %?FIG?
This was decreasing the total flux of the asteroid, as the box used for sumation of the fluxes was hitting mostly sky.
This also caused a ``sawtoothing'' effect, where the lighturve would jump from a minimum to a maximum when the integer coordinate snaped from being a pixel out to back on the asteroid.
This both of these issues can be fixed by calculation of the centre of mass (COM) flux, the average brightness in $xy$ space on a $8\times8$ cut of the image centred on the interpolated position.
I didn't take the highest pixel value in this box because there could be a rouge star, this would drag the COM around, but not by as much as snapping to that highest point.
A \texttt{Photutils} %TODO cite 
aperture photometry call to this COM with a radis of \qty{1.5}{\px} was enough to get fluxes of the same size, but with a far more smooth transition from maximum to minimum, as expected from rotating asteroids.

Not every interpolated point gets a match, due to a variety of reasons, which culminate in there being not an extreme enough of a difference between frames.
A comparison between the two light curves is interesting.

\autoref{Fig:DifFlux} shows these two light curves for a chosen asteroid, ``Ruff''.
This was picked as it has a high number of points matched to detections, but of course still more interpolated points total.
There seems to be a systematic offset between the fluxes, with the interpolated points having consistently lower flux, (here the means differ by 58 counts)
Adjusting aperture to sum the fluxes over to the ``centre of mass'' of the asteroid in each frame does not alleviate this problem.

@@

The completeness of the asteroid population in TESS as a function of their apparent magnitude can be determined.
The apparent magnitude is one of the properties returned from the queries, and is in the V band.
By taking a limit on the if an asteroid is detected \dots %TODO what is the limit
Breaking the average V mag of these objects into \qty{0.5}{\mag} bins for both the sample returned from a query and the sample detected, a ratio of completeness can be calculated per bin.
(Note: If there is no asteroids in a bin, none have a chance of being recovered, so the all did get recovered. I.e. $\frac{0}{0}=1$)
